{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}

### 2017-09-28 sftrends_manu.Rnw
additional formatting away from ES&T

\begin{singlespace}
\title{{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^1$, Thomas W. Jabusch$^2$, Philip R. Trowbridge$^3$, David B. Senn$^4$}}
  \\\\{\textit {\normalsize $^1$Southern California Coastal Water Research Project}}
  \\{\textit {\normalsize 3535 Harbor Blvd, Suite 110, Costa Mesa, CA 92626}}
	\\{\textit {\normalsize Phone: 714-755-3217, Fax: 714-755-3299, Email: \href{mailto:marcusb@sccwrp.org}{marcusb@sccwrp.org}}}
  \\\\{\textit {\normalsize $^2$San Francisco Estuary Institute}}
	\\{\textit {\normalsize 4911 Central Ave, Richmond, CA 94804}}
	\\{\textit {\normalsize Phone: 510-746-7340, Fax: 510-746-7300, Email: \href{mailto:thomasj@sfei.org}{thomasj@sfei.org}}}
	\\\\{\textit {\normalsize $^3$San Francisco Estuary Institute}}
	\\{\textit {\normalsize 4911 Central Ave, Richmond, CA 94804}}
	\\{\textit {\normalsize Phone: 510-746-7345, Fax: 510-746-7300, Email: \href{mailto:philt@sfei.org}{philt@sfei.org}}}
	\\\\{\textit {\normalsize $^4$San Francisco Estuary Institute}}
	\\{\textit {\normalsize 4911 Central Ave, Richmond, CA 94804}}
	\\{\textit {\normalsize Phone: 510-746-7366, Fax: 510-746-7300, Email: \href{mailto:davids@sfei.org}{davids@sfei.org}}}
  \vspace{1in}
  \\ \Sexpr{raw}
	}
\date{}
\maketitle
\end{singlespace}

@@

Trend analysis is a broad discipline that has been applied to time series for the interpretation of environmentally-relevant changes.  Direct evaluation of an observed time series is often insufficient, given that a long-term change can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables\cite{Oneill89,Levin92}.  As a practical approach for water quality evaluation, trend analysis of ecosystem response indicators often focuses on tracking the change in concentrations or loads of nutrients over many years. Response indicators can vary naturally with changing flow conditions and may also reflect long-term effects of management or policy changes. For example, \ac{chla} concentration as a measure of phytoplankton response to nutrient inputs can follow seasonal patterns with cyclical variation in temperature and light changes throughout each year, whereas annual trends can follow long-term variation in nutrient inputs to the system \cite{Cloern96,Cloern10}. Similarly, nutrient trends that vary with hydrologic loading also vary as a function of utilization rates by primary producers or decomposition processes \cite{Sakamoto89,Schultz08,Harding16}. Time series analysis of ecosystem response indicators must simultaneously consider effects of processes at multiple scales and interactions between variables of interest to develop a more comprehensive description of system change.

Appropriate methods for the analysis of change depend largely on the question of interest and characteristics of the environmental dataset.  Trend analyses for aquatic systems have traditionally focused on comparisons between discrete periods of time to estimate direction and magnitude of a trend using non-parametric tests \cite{Hirsch91,Esterby96}. Development of these conventional approaches addressed limitations in historical monitoring datasets related to infrequent sampling and relatively few years of continuous data. Increased availability of multi-decadal datasets, particularly for high profile environments, has accelerated development of trend analysis methods that leverage the descriptive potential of long-term time series from continuous monitoring programs \cite{Bowes09,Halliday12}. These methods are often data-driven where the parameterization of a simple functional model can change smoothly over time.  The \ac{wrtds} approach was developed in this context and has been used to characterize decadal trends in running-water systems \cite{Hirsch10,Sprague11,Medalie12,Hirsch14,Pellerin14,Zhang16}. More recently, the \ac{wrtds} method was adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \cite{Beck15} and the Patuxent River Estuary \cite{Beck17}, and tidally-influenced time series of dissolved oxygen from continuous sonde measurements \cite{Beck15b}. These studies have demonstrated the potential of \ac{wrtds} for trend analysis in tidal waters.

The Sacramento - San Joaquin River Delta (hereafter `Delta') is a mosaic of inflows upstream of the \ac{sfe} that receives and processes inputs from the larger watershed \cite{Jassby00,Jassby02,Jassby08}. Sediment export downstream of the Delta and \ac{wwtp} inputs are primary sources of nutrients for the larger Bay.  Background nutrient concentrations in \ac{sfe} often exceed those associated with excessive primary production, although ecosystem responses symptomatic of eutrophication have historically been infrequent.  Changes in response to additional stressors (e.g., variation in freshwater inputs/withdrawals, invasive species, climate change) suggests that recent conditions have not followed past trajectories and more subtle spatial and temporal variation could provide clues that describe underlying properties of this system \cite{Cloern12b}.  A comprehensive monitoring dataset has been collected at several fixed locations in the upper estuary and Delta for the last four decades \cite{Jabusch16}.  Moreover, nutrient dynamics in the Delta are inherently linked to flow variation from inputs, withdrawal, impoundments, and downstream transport \cite{Novick15}, suggesting that an approach that explicitly considers flow effects is critical for trend analysis.  To date, the regional monitoring dataset for the northern \ac{sfe}, including the Delta, is under-utilized and a comprehensive analysis with \ac{wrtds} could facilitate an understanding of historical and recent changes in water quality.

@@

The Delta region drains a 200 thousand km$^2$ watershed into the \ac{sfe}, which is the largest estuary on the Pacific coast of North America.  The watershed provides water to over 25 million people and irrigation for 18 thousand km$^2$ of agricultural land.  Water enters the \ac{sfe} through the Sacramento and San Joaquin rivers that have a combined inflow of approximately 28 km$^3$ per year, with the Sacramento accounting for 84\% of inflow to the Delta.  The \ac{sfe} system includes the Delta and subembayments of San Francisco Bay (\cref{fig:delt_map}).  Water dynamics in the \ac{sfe} and Delta are governed by inflows from the watershed, tidal exchange with the Pacific Ocean, and water withdrawals for municipal and agricultural use \cite{Jassby00}.  Seasonally, inflows from the watershed peak in the spring and early summer from snowmelt, whereas consumption, withdrawals, and export have steadily increased from 1960 to present, but vary depending on inter-annual climate effects \cite{Cloern12b}. Notable drought periods have occurred from 1976-1977, 1987-1992, and recently from 2013-2015 \cite{Cloern15}.

Orthophosphate (PO$_4^{3-}$) and \ac{din} enter the Delta primarily through the Sacramento and San Joaquin rivers and from municipal \ac{wwtp} inputs.  Annual nutrient export from the Delta region has been estimated as approximately 30 thousand kg d$^{-1}$ of total nitrogen (varying with flow\cite{Novick15}), with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \cite{Jassby08}.  Although nitrogen and phosphorus inputs are considerable, primary production is relatively low and not nutrient-limited \cite{Jassby02,Kimmerer12}.  The resistance of \ac{sfe} to the negative effects of eutrophication has historically been attributed to its unique physical and biological characteristics, including strong tidal mixing that limits stratification in the larger estuary \cite{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks in the northern portion \cite{Thompson08,Crauder16}.  However, recent water quality trends have suggested that resilience to nutrient inputs is decreasing\cite{Lehman05,Cloern07,Lehman10}, which has been attributed to biological invasions \cite{Cohen98} and departures from the historical flow record \cite{Enright09,Cloern12b}, among other factors acting at global scales (e.g., variation in sea surface temperatures).\cite{Cloern07}  The role of nutrients in stimulating primary production in \ac{sfe} has been the focus of several recent investigations.\cite{Dugdale07,Parker12,Glibert14}

@@

Nutrient time series of monthly observations from 1976 to 2013 were obtained for ten sampling stations (\cref{fig:delt_map}, \url{http://water.ca.gov/bdma/meta/Discrete/data.cfm})\cite{IEP13}.  Stations were grouped by location in the study area for comparison: peripheral Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; interior Delta stations D19, D26, D28; and Suisun stations D4, D6, and D7.  These stations were chosen based on continuity of the water quality time series and significance of their geographic location for understanding regional trends.  Time series were complete for all stations except for an approximate ten year gap from 1996-2014 for D19.  Data were minimally processed, with the exception of averaging replicates that occurred on the same day.  The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two). Less than 3\% of all observations were left-censored, although variation was observed between analytes and location.  The ammonium time series had the most censored observations at sites \Sexpr{cvals$Site_Code[1]} (\Sexpr{cvals$cper[1]}\% of all observations), \Sexpr{cvals$Site_Code[2]} (\Sexpr{cvals$cper[2]}\%), \Sexpr{cvals$Site_Code[3]} (\Sexpr{cvals$cper[3]}\%), \Sexpr{cvals$Site_Code[4]} (\Sexpr{cvals$cper[4]}\%), and \Sexpr{cvals$Site_Code[5]} (\Sexpr{cvals$cper[5]}\%).

Daily flow estimates for the Delta region were obtained from the Dayflow software program \cite{IEP16}. The \ac{wrtds} models described below require a matched flow record with the appropriate station to evaluate nutrient trends. Given the complexity of inflows and connectivity of the system, only the inflow estimates from the Sacramento and San Joaquin rivers were used as measures of freshwater influence at each station.  Initial analyses indicated that model fit was not significantly improved with flow estimates from locations closer to each station, nor was model fit improved using lagged times series.  As such, the Sacramento daily flow time series was used to account for flow effects at C3, D19, D26, D28, and MD10, and the San Joaquin time series was used for C10 and P8 based on station proximity to each inflow.  Salinity observations at D4, D6, and D7 in Suisun Bay were used as more appropriate measures of freshwater variation, given the stronger tidal influence at these stations.  Salinity has been used as a tracer of freshwater influence for the application of \ac{wrtds} models in tidal waters \cite{Beck15}.

@@

A total of thirty \ac{wrtds} models were created, one for each nitrogen analyte at each station.  The functional form of \ac{wrtds} is a simple regression\cite{Hirsch10} that models the log-transformed response variable as a function of time, flow, and season:

@@

\noindent where $N$ is one of three nitrogen analytes, time $t$ is a continuous variable as decimal time to capture the annual or seasonal trend, and $Q$ is the flow variable (either flow or salinity depending on station).  The \ac{wrtds} model is a moving window regression that fits unique parameters at each observation point in the time series.  Models applied herein were based on a tidal adaptation of the original method\cite{Beck15} and were fit to describe the conditional mean response using a weighted Tobit model for left-censored data \cite{Tobin58}. All analyses used the WRTDStidal package written by the authors for the R statistical programming language \cite{Beck16b,RDCT17}.

A hallmark of the \ac{wrtds} approach is the description of flow-normalized trends that are independent of variation from freshwater inflows\cite{Hirsch10}. Flow-normalized trends for each analyte at each station were used to describe long-term changes in different annual and seasonal periods.  Specifically, flow-normalized trends in each analyte were summarized as both medians and percent changes from the beginning to end of annual groupings from 1976-1995 and 1996-2013, and seasonal groupings of March-April-May (spring), June-July-August (summer), September-October-November (fall), and December-January-February (winter) within each annual grouping. These annual and seasonal groupings were chosen for continuity with similar comparisons in Ref.~\citenum{Jabusch16} and as approximate twenty year midpoints in the time series.

Trends in each annual and seasonal grouping were based on seasonal Kendall tests of the flow-normalized predictions. This test is a modification of the non-parametric Kendall test that accounts for variation across seasons in the response variable \cite{Hirsch82,Millard13}.  Results from the test can be used to evaluate the direction, magnitude, and significance of a monotonic change within the period of observation.  The estimated rate of change per year is also returned as the Theil-Sen slope and was interpreted as the percent change per year when divided by the median value of the response variable in the period of observation.\cite{Jassby08}  Trends in annual groupings were based on all monthly observations within relevant years, whereas seasonal groupings were based only on the relevant months across years.  Seasonal Kendall tests were also used to describe trends in the observed data.  These trends were compared with those based on the flow-normalized trends to evaluate the improved ability of \ac{wrtds} to describe trends that are independent of flow.

\section{Results and Discussion}

@@

The observed time series for the ten Delta - Suisun Bay stations had substantial variation in scale among the nitrogen analytes and differences in apparent seasonal trends (\cref{fig:obsdat}).  \ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.  However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D26) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower\cite{Novick15}.  By location, observed concentrations of \ac{din} for the entire time series were higher on average for the peripheral stations (C3, C10, MD10, P8; mean $\pm$ s.e.: \Sexpr{scinot(obslo$Delta['mean'])}$\pm$\Sexpr{scinot(obslo$Delta['se'])} \mgl) and similar for the interior (D19, D26, D28, \Sexpr{scinot(obslo$Middle['mean'])}$\pm$\Sexpr{scinot(obslo$Middle['se'], 3)}) and Suisun Bay stations (D4, D6, D7, \Sexpr{scinot(obslo$Suisun['mean'])}$\pm$\Sexpr{scinot(obslo$Suisun['se'], 3)}).  Average concentrations were highest at \Sexpr{obssi$din[1, 'Site_Code']} (\Sexpr{scinot(obssi$din[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[1, 'se'])} \mgl) and lowest at \Sexpr{obssi$din[2, 'Site_Code']} (\Sexpr{scinot(obssi$din[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[2, 'se'], 3)}) for \ac{din}, highest at \Sexpr{obssi$nh[1, 'Site_Code']} (\Sexpr{scinot(obssi$nh[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[1, 'se'])}) and lowest at \Sexpr{obssi$nh[2, 'Site_Code']} (\Sexpr{scinot(obssi$nh[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[2, 'se'], 3, 3)}) for ammonium, and highest at \Sexpr{obssi$no23[1, 'Site_Code']} (\Sexpr{scinot(obssi$no23[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[1, 'se'])}) and lowest at \Sexpr{obssi$no23[2, 'Site_Code']} (\Sexpr{scinot(obssi$no23[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[2, 'se'], 3, 3)}) for nitrite/nitrate. Mean observed concentrations were also higher later in the time series for all analytes.  For example, average \ac{din} across all stations was \Sexpr{scinot(obsyr$din[1, 'mean'])}$\pm$\Sexpr{scinot(obsyr$din[1, 'se'])} \mgl for \Sexpr{obsyr$din[1, 'yr']}, compared to \Sexpr{scinot(obsyr$din[2, 'mean'])}$\pm$\Sexpr{scinot(obsyr$din[2, 'se'])} for \Sexpr{obsyr$din[2, 'yr']}. Seasonal changes across all years showed that nitrogen concentrations were generally lower in the summer and higher in the winter, although observed patterns were inconsistent between sites.  For example, site MD10 had distinct seasonal spikes for elevated \ac{din} in the winter, whereas other stations had less prominent seasonal maxima (e.g., C3, D7, \cref{fig:obsdat}).

@@

Estimated trends from Seasonal Kendall tests on the observed data varied considerably between sites and analytes (\cref{fig:trndcomp1}). Significant trends were observed from 1976-1995 for \Sexpr{obsdir$din[1, 5]} of ten sites for \ac{din} (\Sexpr{obsdir$din[1, 3]} increasing, \Sexpr{obsdir$din[1, 4]} decreasing), \Sexpr{obsdir$nh[1, 5]} sites for ammonium (\Sexpr{obsdir$nh[1, 3]} increasing, \Sexpr{obsdir$nh[1, 4]} decreasing), and \Sexpr{obsdir$no23[1, 5]} sites for nitrite/nitrate (\Sexpr{obsdir$no23[1, 3]} increasing, \Sexpr{obsdir$no23[1, 4]} decreasing).  Decreasing trends were more common for the observed data from 1996-2013.  \Sexpr{cap(obsdir$din[2, 5])} sites had significant trends for \ac{din} (\Sexpr{obsdir$din[2, 3]} increasing, \Sexpr{obsdir$din[2, 4]} decreasing), \Sexpr{obsdir$nh[2, 5]} sites for ammonium (\Sexpr{obsdir$nh[2, 3]} increasing, \Sexpr{obsdir$nh[2, 4]} decreasing), and \Sexpr{obsdir$no23[2, 5]} sites for nitrite/nitrate (\Sexpr{obsdir$no23[2, 3]} increasing, \Sexpr{obsdir$no23[2, 4]} decreasing). P8 had a relatively large decrease in ammonium (\Sexpr{scinot(P8dec[2, 'perchg'], 1, 1)}\% change per year) for the second annual period compared to all other sites (see next section). Trends by season were similar such that increases were generally observed in all seasons from 1976-1995 (Figure S1) and decreases were observed for 1996-2013 (Figure S2).  Trends for the seasonal comparisons were noisier and significant changes were less common compared to the annual comparisons.

@@

>>>>

\begin{singlespace}
\title{{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^{1*}$, Thomas W. Jabusch$^2$, Philip R. Trowbridge$^2$, David B. Senn$^2$}}
  \\\\{\textit {\normalsize $^*$Corresponding author: \href{mailto:marcusb@sccwrp.org}{marcusb@sccwrp.org}}}
  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
  \\\\{\textit {\normalsize Current address: Southern California Coastal Water Research Project}}
  \\{\textit {\normalsize 3535 Harbor Blvd, Suite 110, Costa Mesa, CA 92626}}
  \\\\{\textit {\normalsize $^2$San Francisco Estuary Institute}}
	\\{\textit {\normalsize 4911 Central Ave, Richmond, CA 94804}}
  \vspace{1in}
  \\ \Sexpr{raw}
	}
\date{}
\maketitle
\end{singlespace}
\clearpage

@@

Trend analysis is a broad discipline that has been applied to time series for the interpretation of environmentally-relevant changes.  Direct evaluation of an observed time series is often insufficient, given that a long-term change can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables \citep{Oneill89,Levin92}.  As a practical approach for water quality evaluation, trend analysis of ecosystem response indicators often focuses on tracking the change in concentrations or loads of nutrients over many years. Response indicators can vary naturally with changing flow conditions and may also reflect long-term effects of management or policy changes. For example, \ac{chla} concentration as a measure of phytoplankton response to nutrient inputs can follow seasonal patterns with cyclical variation in temperature and light changes throughout each year, whereas annual trends can follow long-term variation in nutrient inputs to the system \citep{Cloern96,Cloern10}. Similarly, nutrient trends that vary with hydrologic loading also vary as a function of utilization rates by primary producers or decomposition processes \citep{Sakamoto89,Schultz08,Harding16}. Time series analysis of ecosystem response indicators must simultaneously consider effects of processes at multiple scales and interactions between variables of interest to develop a more comprehensive description of system change.

Appropriate methods for the analysis of change depend largely on the question of interest and characteristics of the environmental dataset.  Trend analyses for aquatic systems have traditionally focused on comparisons between discrete periods of time to estimate direction and magnitude of a trend using non-parametric tests \citep{Hirsch91,Esterby96}. Development of these conventional approaches addressed limitations in historical monitoring datasets related to infrequent sampling and relatively few years of continuous data. Increased availability of multi-decadal datasets, particularly for high profile environments, has accelerated development of trend analysis methods that leverage the descriptive potential of long-term time series from continuous monitoring programs \citep{Bowes09,Halliday12}. These methods are often data-driven where the parameterization of a simple functional model can change smoothly over time.  The \ac{wrtds} approach was developed in this context and has been used to characterize decadal trends in running-water systems \citep{Hirsch10,Sprague11,Medalie12,Hirsch14,Pellerin14,Zhang16}. More recently, the \ac{wrtds} method was adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \citep{Beck15} and the Patuxent River Estuary \citep{Beck17}, and tidally-influenced time series of dissolved oxygen from continuous sonde measurements \citep{Beck15b}. These studies have demonstrated the potential of \ac{wrtds} for trend analysis in tidal waters.

The Sacramento - San Joaquin River Delta (hereafter `Delta') is a mosaic of inflows upstream of the \ac{sfe} that receives and processes inputs from the larger watershed \citep{Jassby00,Jassby02,Jassby08}. Sediment export downstream of the Delta and \ac{wwtp} inputs are primary sources of nutrients for the larger Bay.  Background nutrient concentrations in \ac{sfe} often exceed those associated with excessive primary production, although ecosystem responses symptomatic of eutrophication have historically been infrequent.  Changes in response to additional stressors (e.g., variation in freshwater inputs/withdrawals, invasive species, climate change) suggests that recent conditions have not followed past trajectories and more subtle spatial and temporal variation could provide clues that describe underlying properties of this system \citep{Cloern12b}.  A comprehensive monitoring dataset has been collected at several fixed locations in the upper estuary and Delta for the last four decades \citep{Jabusch16}.  Moreover, nutrient dynamics in the Delta are inherently linked to flow variation from inputs, withdrawal, impoundments, and downstream transport \citep{Novick15}, suggesting that an approach that explicitly considers flow effects is critical for trend analysis.  To date, the regional monitoring dataset for the northern \ac{sfe}, including the Delta, is under-utilized and a comprehensive analysis with \ac{wrtds} could facilitate an understanding of historical and recent changes in water quality.

@@

The Delta region drains a 200 thousand km$^2$ watershed into the \ac{sfe}, which is the largest estuary on the Pacific coast of North America.  The watershed provides water to over 25 million people and irrigation for 18 thousand km$^2$ of agricultural land.  Water enters the \ac{sfe} through the Sacramento and San Joaquin rivers that have a combined inflow of approximately 28 km$^3$ per year, with the Sacramento accounting for 84\% of inflow to the Delta.  The \ac{sfe} system includes the Delta and subembayments of San Francisco Bay (\cref{fig:delt_map}).  Water dynamics in the \ac{sfe} and Delta are governed by inflows from the watershed, tidal exchange with the Pacific Ocean, and water withdrawals for municipal and agricultural use \citep{Jassby00}.  Seasonally, inflows from the watershed peak in the spring and early summer from snowmelt, whereas consumption, withdrawals, and export have steadily increased from 1960 to present, but vary depending on inter-annual climate effects \citep{Cloern12b}. Notable drought periods have occurred from 1976-1977, 1987-1992, and recently from 2013-2015 \citep{Cloern15}.

Orthophosphate (PO$_4^{3-}$) and \ac{din} enter the Delta primarily through the Sacramento and San Joaquin rivers and from municipal \ac{wwtp} inputs.  Annual nutrient export from the Delta region has been estimated as approximately 30 thousand kg d$^{-1}$ of total nitrogen (varying with flow\citep{Novick15}), with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \citep{Jassby08}.  Although nitrogen and phosphorus inputs are considerable, primary production is relatively low and not nutrient-limited \citep{Jassby02,Kimmerer12}.  The resistance of \ac{sfe} to the negative effects of eutrophication has historically been attributed to its unique physical and biological characteristics, including strong tidal mixing that limits stratification in the larger estuary \citep{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks in the northern portion \citep{Thompson08,Crauder16}.  However, recent water quality trends have suggested that resilience to nutrient inputs is decreasing\citep{Lehman05,Cloern07,Lehman10}, which has been attributed to biological invasions \citep{Cohen98} and departures from the historical flow record \citep{Enright09,Cloern12b}, among other factors acting at global scales (e.g., variation in sea surface temperatures, \cite{Cloern07})  The role of nutrients in stimulating primary production in \ac{sfe} has been the focus of several recent investigations\ \citep{Dugdale07,Parker12,Glibert14}.

Quantitative descriptions of nutrient dynamics in the Delta are challenging given multiple sources and the volume of water that is exchanged with natural and anthropogenic processes.  A comprehensive evaluation using mass-balance models to describe nutrient dynamics in the Delta demonstrated that nitrogen enters the system in different forms and is processed at different rates before export or removal \citep{Novick15}. For example, a majority of ammonium entering the system during the summer is nitrified or assimilated, whereas a considerable percentage of total nitrogen load to the Delta is exported.  Although, the focus of our analysis is not to quantify sources or sinks of nitrogen species, a quantitative evaluation of long-term trends will provide a more comprehensive historical interpretation to hypothesize the effects of future changes in the context of known dynamics.  Nutrients in the Delta also vary with seasonal and annual changes in the delivery of water inflows and water exports directly from the system \citep{Jassby00,Jassby08}.  Our analysis explicitly accounts for the effects of flow changes on nutrient response to better understand variation both within the Delta and potential mechanisms of downstream tranport.

@@

Nutrient time series of monthly observations from 1976 to 2013 were obtained for ten sampling stations (\cref{fig:delt_map}, \url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}, \cite{IEP13}).  Stations were grouped by location in the study area for comparison: peripheral Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; interior Delta stations D19, D26, D28; and Suisun stations D4, D6, and D7.  These stations were chosen based on continuity of the water quality time series and significance of their geographic location for understanding regional trends.  Time series were complete for all stations except for an approximate ten year gap from 1996-2014 for D19.  Data were minimally processed, with the exception of averaging replicates that occurred on the same day.  The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two). Less than 3\% of all observations were left-censored, although variation was observed between analytes and location.  The ammonium time series had the most censored observations at sites \Sexpr{cvals$Site_Code[1]} (\Sexpr{cvals$cper[1]}\% of all observations), \Sexpr{cvals$Site_Code[2]} (\Sexpr{cvals$cper[2]}\%), \Sexpr{cvals$Site_Code[3]} (\Sexpr{cvals$cper[3]}\%), \Sexpr{cvals$Site_Code[4]} (\Sexpr{cvals$cper[4]}\%), and \Sexpr{cvals$Site_Code[5]} (\Sexpr{cvals$cper[5]}\%).

Daily flow estimates for the Delta region were obtained from the Dayflow software program \citep{IEP16}. The \ac{wrtds} models described below require a matched flow record with the appropriate station to evaluate nutrient trends. Given the complexity of inflows and connectivity of the system, only the inflow estimates from the Sacramento and San Joaquin rivers were used as measures of freshwater influence at each station.  Initial analyses indicated that model fit was not significantly improved with flow estimates from locations closer to each station, nor was model fit improved using lagged times series.  As such, the Sacramento daily flow time series was used to account for flow effects at C3, D19, D26, D28, and MD10, and the San Joaquin time series was used for C10 and P8 based on station proximity to each inflow.  Salinity observations at D4, D6, and D7 in Suisun Bay were used as more appropriate measures of freshwater variation, given the stronger tidal influence at these stations.  Salinity has been used as a tracer of freshwater influence for the application of \ac{wrtds} models in tidal waters \citep{Beck15}.

@@

A total of thirty \ac{wrtds} models were created, one for each nitrogen analyte at each station.  The functional form of \ac{wrtds} is a simple regression \citep{Hirsch10} that models the log-transformed response variable as a function of time, flow, and season:

@@

\noindent where $N$ is one of three nitrogen analytes, time $t$ is a continuous variable as decimal time to capture the annual or seasonal trend, and $Q$ is the flow variable (either flow or salinity depending on station).  The \ac{wrtds} model is a moving window regression that fits unique parameters at each observation point in the time series.  Models applied herein were based on a tidal adaptation of the original method \citep{Beck15} and were fit to describe the conditional mean response using a weighted Tobit model for left-censored data \citep{Tobin58}. All analyses used the WRTDStidal package written by the authors for the R statistical programming language \citep{Beck16b,RDCT17}.

A hallmark of the \ac{wrtds} approach is the description of flow-normalized trends that are independent of variation from freshwater inflows \citep{Hirsch10}. Flow-normalized trends for each analyte at each station were used to describe long-term changes in different annual and seasonal periods.  Specifically, flow-normalized trends in each analyte were summarized as both medians and percent changes from the beginning to end of annual groupings from 1976-1995 and 1996-2013, and seasonal groupings of March-April-May (spring), June-July-August (summer), September-October-November (fall), and December-January-February (winter) within each annual grouping. These annual and seasonal groupings were chosen for continuity with similar comparisons in \citet{Jabusch16} and as approximate twenty year midpoints in the time series.

Trends in each annual and seasonal grouping were based on seasonal Kendall tests of the flow-normalized predictions. This test is a modification of the non-parametric Kendall test that accounts for variation across seasons in the response variable \citep{Hirsch82,Millard13}.  Results from the test can be used to evaluate the direction, magnitude, and significance of a monotonic change within the period of observation.  The estimated rate of change per year is also returned as the Theil-Sen slope and was interpreted as the percent change per year when divided by the median value of the response variable in the period of observation \citep{Jassby08}.  Trends in annual groupings were based on all monthly observations within relevant years, whereas seasonal groupings were based only on the relevant months across years.  Seasonal Kendall tests were also used to describe trends in the observed data.  These trends were compared with those based on the flow-normalized trends to evaluate the improved ability of \ac{wrtds} to describe trends that are independent of flow.

\section{Results}

@@

The observed time series for the ten Delta - Suisun Bay stations had substantial variation in scale among the nitrogen analytes and differences in apparent seasonal trends (\cref{fig:obsdat}).  \ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.  However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D26) had higher concentrations of ammonium during winter months when phytoplankton assimilation is lower \citep{Novick15}.  By location, observed concentrations of \ac{din} for the entire time series were higher on average for the peripheral stations (C3, C10, MD10, P8; mean $\pm$ s.e.: \Sexpr{scinot(obslo$Delta['mean'])}$\pm$\Sexpr{scinot(obslo$Delta['se'])} \mgl) and similar for the interior (D19, D26, D28, \Sexpr{scinot(obslo$Middle['mean'])}$\pm$\Sexpr{scinot(obslo$Middle['se'], 3)}) and Suisun Bay stations (D4, D6, D7, \Sexpr{scinot(obslo$Suisun['mean'])}$\pm$\Sexpr{scinot(obslo$Suisun['se'], 3)}).  Average concentrations were highest at \Sexpr{obssi$din[1, 'Site_Code']} (\Sexpr{scinot(obssi$din[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[1, 'se'])} \mgl) and lowest at \Sexpr{obssi$din[2, 'Site_Code']} (\Sexpr{scinot(obssi$din[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[2, 'se'], 3)}) for \ac{din}, highest at \Sexpr{obssi$nh[1, 'Site_Code']} (\Sexpr{scinot(obssi$nh[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[1, 'se'])}) and lowest at \Sexpr{obssi$nh[2, 'Site_Code']} (\Sexpr{scinot(obssi$nh[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[2, 'se'], 3, 3)}) for ammonium, and highest at \Sexpr{obssi$no23[1, 'Site_Code']} (\Sexpr{scinot(obssi$no23[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[1, 'se'])}) and lowest at \Sexpr{obssi$no23[2, 'Site_Code']} (\Sexpr{scinot(obssi$no23[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[2, 'se'], 3, 3)}) for nitrite/nitrate. Mean observed concentrations were also higher later in the time series for all analytes.  For example, average \ac{din} across all stations was \Sexpr{scinot(obsyr$din[1, 'mean'])}$\pm$\Sexpr{scinot(obsyr$din[1, 'se'])} \mgl for \Sexpr{obsyr$din[1, 'yr']}, compared to \Sexpr{scinot(obsyr$din[2, 'mean'])}$\pm$\Sexpr{scinot(obsyr$din[2, 'se'])} for \Sexpr{obsyr$din[2, 'yr']}. Seasonal changes across all years showed that nitrogen concentrations were generally lower in the summer and higher in the winter, although observed patterns were inconsistent between sites.  For example, site MD10 had distinct seasonal spikes for elevated \ac{din} in the winter, whereas other stations had less prominent seasonal maxima (e.g., C3, D7, \cref{fig:obsdat}).

@@

Estimated trends from Seasonal Kendall tests on the observed data varied considerably between sites and analytes (\cref{fig:trndcomp1}). Significant trends were observed from 1976-1995 for \Sexpr{obsdir$din[1, 5]} of ten sites for \ac{din} (\Sexpr{obsdir$din[1, 3]} increasing, \Sexpr{obsdir$din[1, 4]} decreasing), \Sexpr{obsdir$nh[1, 5]} sites for ammonium (\Sexpr{obsdir$nh[1, 3]} increasing, \Sexpr{obsdir$nh[1, 4]} decreasing), and \Sexpr{obsdir$no23[1, 5]} sites for nitrite/nitrate (\Sexpr{obsdir$no23[1, 3]} increasing, \Sexpr{obsdir$no23[1, 4]} decreasing).  Decreasing trends were more common for the observed data from 1996-2013.  \Sexpr{cap(obsdir$din[2, 5])} sites had significant trends for \ac{din} (\Sexpr{obsdir$din[2, 3]} increasing, \Sexpr{obsdir$din[2, 4]} decreasing), \Sexpr{obsdir$nh[2, 5]} sites for ammonium (\Sexpr{obsdir$nh[2, 3]} increasing, \Sexpr{obsdir$nh[2, 4]} decreasing), and \Sexpr{obsdir$no23[2, 5]} sites for nitrite/nitrate (\Sexpr{obsdir$no23[2, 3]} increasing, \Sexpr{obsdir$no23[2, 4]} decreasing). P8 had a relatively large decrease in ammonium (\Sexpr{scinot(P8dec[2, 'perchg'], 1, 1)}\% change per year) for the second annual period compared to all other sites (see next section). Trends by season were similar such that increases were generally observed in all seasons from 1976-1995 (\cref{fig:trndcomp2}) and decreases were observed for 1996-2013 (\cref{fig:trndcomp3}).  Trends for the seasonal comparisons were noisier and significant changes were less common compared to the annual comparisons.

@@

Two stations were chosen to demonstrate use of \ac{wrtds} to develop a more comprehensive description of decadal trends in the Delta.  The selected case studies focused on \begin{inparaenum}[1\upshape)]		
\item effects of wastewater treatment upgrades upstream of P8, and		
\item effects of biological invasion on nutrient dynamics in Suisun Bay using observations from D7.		
\end{inparaenum}		
Each case study is built around hypotheses that results from \ac{wrtds} models were expected to support, both as a general description and for additional testing with alternative methods.
